A Decision Procedure for (Co)datatypes in SMT Solvers

### 2016-02-09 jour/jour.tex
more

Datatypes and codatatypes share many basic properties. All properties
below are implicitly universally quantified and range over all $i$, $j$,
$\jPrime$, and $k$ within bounds:
% and to all possible splits of the
%$n$-ary constructor $\const{C}_{i\negvthinspace j}$'s argument list into $\bar x,
%y, \bar z$:
%
\[
\begin{aligned}[t]
\text{Distinctness:}\quad
  & %\forall \bar x, \bar y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(\bar x) \tneq \const{C}_{i\negvthinspace \jPrime}(\bar y) \quad\text{if $j \not= \jPrime$}}
  \\[-0.5\jot]
\text{Injectivity:}\quad
  & %\forall x_1,\ldots,x_{n_{i\negvthinspace j}} y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}}) \teq \const{C}_{i\negvthinspace j}(y_1,\ldots,y_{n_{i\negvthinspace j}}) \medrightarrow x_k \teq y_k}
  \\[-0.5\jot]
\text{Exhaustiveness:}\quad
  & \smash{\const{d}_{i1}(x) \mathrel\lor \cdots \mathrel\lor \const{d}_{im_i}(x)}
  \\[-0.5\jot]
\text{Selection:}\quad
  & \smash{\const{s}_{i\negvthinspace j}^{\,k}(\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}})) \teq x_k}
\end{aligned}
\]
%
Expressed in the algebraic jargon, exhaustiveness helps ensure that ``no
junk'' exists, whereas distinctness and injectivity guarantee that ``no
confusion'' can arise.
The result of selectors applied to the wrong
constructor is left completely unspecified.
%
%
%  * how to deal with "wrong" selectors, e.g.
%        hd(nil1) = hd(nil2)?
%      * leave them unspecified; hence hd(nil1) = hd(nil2) in some models,
%        and not in other models
%
Datatypes are additionally characterized by an induction principle for proving
a conjunction of properties $P_1,\ldots,P_{\ell}$ over arbitrary
values $v_1 : \delta_1, \ldots, v_\ell : \delta_\ell$:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
\AXC{\strut$\bigwedge_{\vthinspace i,j}\, \forall x_1 \ldots \vthinspace x_{n_{i\negvthinspace j}}.\; \bigl(\bigwedge_{\vthinspace k}\, \mathcalx{IH}[x_k]\bigr) \medrightarrow P_i[\const C_{i\negvthinspace j}(x_1,\ldots,x_{n_{ij}})]$}
\UIC{\strut$\bigwedge_{\vthinspace i}\, P_i[v_i]$}
\DP
\end{aligned}
\]
The notation $\mathcalx{IH}[x]$
denotes either $P_{\iPrime}[x]$ if there exists some $\iPrime$ such that
the formula is type-correct or else $\top$ (truth).
The induction principle ensures that the interpretation of datatypes is standard.
For the natural numbers constructed from $\const{Z}$ and $\const{S}$,
induction prohibits models that contain cyclic values---e.g.,
an $n$ such that $n \teq \const{S}(n)$---or even infinite acyclic values
$\const{S}(\const{S}(\ldots))$.

For codatatypes, the dual notion is called coinduction. It makes it possible to
derive the equality of pairs of codatatype values
$v_1, w_1 \mathbin: \delta_1,\,\ldots,\,v_\ell, w_\ell \mathbin: \delta_\ell$, based on
suitable coinduction witnesses $R_1,\ldots,R_\ell$:
%
\[
\begin{aligned}[t]
\text{Coinduction:}\quad
\AXC{\strut$\begin{gathered}\textstyle \smash{\bigwedge_{\vthinspace i}\, R_i[v_i, w_i]} \\[-\jot]\textstyle
\bigwedge_{\vthinspace i}\, \forall v\; w.\;\, R_i[v, w] \medrightarrow
  \bigwedge_{\vthinspace j}\, \const{d}_{\negvthinspace j}(v) \mathrel\land \const{d}_{\negvthinspace j}(w)
  \medrightarrow \bigwedge_{\vthinspace k}\, \const{s}_{i\negvthinspace j}^{\,k}(v) \sim \const{s}_{i\negvthinspace j}^{\,k}(w)
\end{gathered}$}
\UIC{\strut$\bigwedge_{\vthinspace i}\, v_i \teq w_i$}
\DP
\end{aligned}
\]
The notation $x \sim y$ stands for $R_{\iPrime}[x, y]$ if there exists some
$\iPrime$ such that the formula is type-correct or $x \teq y$ otherwise.
The second premise ensures that the coinduction witnesses are
bisimulations. The first premise and the conclusion capture the notion that
equality is the largest bisimulation on codatatypes. Thus, the coinduction
principle encodes a form of extensionality:\ Two values that yield the same
observations must be equal, where the observations are made through selectors
and discriminators.

@@

Formally, $\mu$-\emph{terms} are defined recursively as being either a variable $x$
or an applied constructor
$\MU x.\: \const{C}( \vec t\, )$ for some $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ of the expected types.
The variable~$x$ need not occur free in the $\mu$-binder's body, in which case
the binder can be omitted.
$\FV( t )$ denotes the set of free variables occurring in the $\mu$-term~$t$.
A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains a bound variable. The $\alpha$\emph{-equiv\-alence} relation $t \aequiv u$
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$,
but
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{D}( y, x ) \not\aequiv \MU x.\; \const{D}( z, x )
\not\aequiv \MU y.\; \const{D}( y, x )$.
Two $\mu$-terms can denote the same value despite being $\alpha$-disequivalent---e.g.,
$\MU x.\; \const{S}( x ) \not\aequiv \MU y.\; \const{S}( \const{S}( y ) )$.
%In the \rn{Acyclic} rule,

@@

However, the presence of quantifiers poses an additional challenge in
the context of (co)datatypes. Quantified formulas can entail an upper bound on
the cardinality of an uninterpreted type $\ty{u}$.
By assuming that uninterpreted types have infinite cardinality, the calculus
presented in Section~\ref{sec:the-decision-procedure} is \relax{incomplete}
since it may fail to recognize cases where \rn{Split} and \rn{Single} should be
applied.
This does not %necessarily
impact the correctness of the procedure in this setting,
since %by assumption
the solver is already incomplete for quantified formulas.

Nonetheless, two techniques help increase the precision of the solver.
First, we can apply \rn{Split} to datatype terms whose cardinality depends on the finiteness of
uninterpreted types.
Second, we can conditionally apply \rn{Single} to codatatype terms that may have cardinality one.
For example, the codatatype $\ty{stream}_{\vthinspace\ty{u}}$ %, where $\ty{u}$ is uninterpreted.
has cardinality one precisely when $\ty{u}$ has cardinality one.
If there exist two equivalence classes $\ec{s}$ and $\ec{t}$ for this type,
the implementation adds the clause %of the form
$(\exists x\;y\typ{\ty{u}}.\;\, x \not\teq y ) \mathrel\vee s \teq t$ to $F\!$,
which states that either the cardinality of $\ty{u}$ is greater than one or
$s$ must be equal to $t$.

@@

The decision procedure for (co)datatypes is useful both for proving (via
negation, in the refutational style) and for model finding
\cite{reynolds-et-al-2013,ge-de-moura-2009}.
It is in fact vital for finite model finding,
because the acyclicity and uniqueness rules are necessary for solution
soundness, without which the generated models would often be
spurious. For example, given the axioms
\begin{align*}
\const{zeros} & = \const{SCons}(0,\:\const{zeros})
& \const{sconst}~n & = \const{SCons}(n,\:\const{sconst}~n)
\end{align*}
on streams, the conjecture
$\const{zeros} = \const{sconst}~0$
would be ``refuted'' by a spurious countermodel that interprets
\const{zero} and $\const{sconst}~0$ by two distinct
values that both correspond to the term
$\MU s.\; \const{SCons}(0, s)$,
violating uniqueness.

>>>>

Datatypes and codatatypes share many basic properties. All properties
below are implicitly universally quantified and range over all $i$, $j$,
$\jPrime$, and $k$ within bounds:
% and to all possible splits of the
%$n$-ary constructor $\const{C}_{i\negvthinspace j}$'s argument list into $\bar x,
%y, \bar z$:
%
\[
\begin{aligned}[t]
\text{Distinctness:}\quad
  & %\forall \bar x, \bar y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(\bar x) \tneq \const{C}_{i\negvthinspace \jPrime}(\bar y) \quad\text{if $j \not= \jPrime$}}
  \\[-0.5\jot]
\text{Injectivity:}\quad
  & %\forall x_1,\ldots,x_{n_{i\negvthinspace j}} y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}}) \teq \const{C}_{i\negvthinspace j}(y_1,\ldots,y_{n_{i\negvthinspace j}}) \medrightarrow x_k \teq y_k}
  \\[-0.5\jot]
\text{Exhaustiveness:}\quad
  & \smash{\const{d}_{i1}(x) \mathrel\lor \cdots \mathrel\lor \const{d}_{im_i}(x)}
  \\[-0.5\jot]
\text{Selection:}\quad
  & \smash{\const{s}_{i\negvthinspace j}^{\,k}(\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}})) \teq x_k}
\end{aligned}
\]
%
Expressed in the algebraic jargon, exhaustiveness helps ensure that ``no
junk'' exists, whereas distinctness and injectivity guarantee that ``no
confusion'' can arise.
The result of selectors applied to the wrong
constructor is left completely unspecified.
%
%
%  * how to deal with "wrong" selectors, e.g.
%        hd(nil1) = hd(nil2)?
%      * leave them unspecified; hence hd(nil1) = hd(nil2) in some models,
%        and not in other models
%
Datatypes are additionally characterized by an induction principle for proving
a conjunction of properties $P_1,\ldots,P_{\ell}$ over arbitrary
values $v_1 : \delta_1, \ldots, v_\ell : \delta_\ell$:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
\AXC{\strut$\bigwedge_{\vthinspace i=1}^\ell \bigwedge_{\vthinspace j=1}^{m_i}\, \forall x_1 \ldots \vthinspace x_{n_{i\negvthinspace j}}.\; \bigl(\bigwedge_{\vthinspace k=1}^{ n_{i\negvthinspace j}}\, \mathcalx{IH}[x_k]\bigr) \medrightarrow P_i[\const C_{i\negvthinspace j}(x_1,\ldots,x_{n_{ij}})]$}
\UIC{\strut$\bigwedge_{\vthinspace i=1}^{\ell}\, P_i[v_i]$}
\DP
\end{aligned}
\]
The notation $\mathcalx{IH}[x]$
denotes either $P_{\iPrime}[x]$ if there exists some $\iPrime$ such that
the formula is type-correct or else $\top$ (truth).
The induction principle ensures that the interpretation of datatypes is standard.
For the natural numbers constructed from $\const{Z}$ and $\const{S}$,
induction prohibits models that contain cyclic values---e.g.,
an $n$ such that $n \teq \const{S}(n)$---or even infinite acyclic values
$\const{S}(\const{S}(\ldots))$.

For codatatypes, the dual notion is called coinduction. It makes it possible to
derive the equality of pairs of codatatype values
$v_1, w_1 \mathbin: \delta_1,\,\ldots,\,v_\ell, w_\ell \mathbin: \delta_\ell$, based on
suitable coinduction witnesses $R_1,\ldots,R_\ell$:
%
\[
\begin{aligned}[t]
\text{Coinduction:}\quad
\AXC{\strut$\begin{gathered}\textstyle \smash{\bigwedge_{\vthinspace i=1}^\ell \, R_i[v_i, w_i]} \\[-\jot]\textstyle
\bigwedge_{\vthinspace i=1}^\ell\, \bigl(\forall v\; w.\;\, R_i[v, w] \medrightarrow
  \bigwedge_{\vthinspace j=1}^{m_i}\, \bigl(\const{d}_{\negvthinspace j}(v) \mathrel\land \const{d}_{\negvthinspace j}(w)
  \medrightarrow \bigwedge_{\vthinspace k=1}^{n_{i\negvthinspace j}}\, \const{s}_{i\negvthinspace j}^{\,k}(v) \sim \const{s}_{i\negvthinspace j}^{\,k}(w)\bigr)\bigr)
\end{gathered}$}
\UIC{\strut$\bigwedge_{\vthinspace i=1}^\ell \, v_i \teq w_i$}
\DP
\end{aligned}
\]
The notation $x \sim y$ stands for $R_{\iPrime}[x, y]$ if there exists some
$\iPrime$ such that the formula is type-correct or $x \teq y$ otherwise.
The second premise ensures that the coinduction witnesses are
bisimulations. The first premise and the conclusion capture the notion that
equality is the largest bisimulation on codatatypes. Thus, the coinduction
principle encodes a form of extensionality:\ Two values that yield the same
observations must be equal, where the observations are made through selectors
and discriminators.

@@

Formally, $\mu$-\emph{terms} are defined recursively as being either a variable $x$
or an applied constructor
$\MU x.\: \const{C}( \vec t\, )$ for some $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ of the expected types.
The variable~$x$ need not occur free in the $\mu$-binder's body, in which case
the binder can be omitted.
$\FV( t )$ denotes the set of free variables occurring in the $\mu$-term~$t$.
A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains a bound variable. The $\alpha$\emph{-equiv\-alence} relation $t \aequiv u$
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$,
but
$\MU x.\; \const{C}( x )$, $\MU x.\; \const{D}( y, x )$, $\MU x.\; \const{D}( z, x )$, and
$\MU y.\; \const{D}( y, x )$ are all $\alpha$-disequivalent.
Two $\mu$-terms can denote the same value despite being $\alpha$-disequivalent---e.g.,
$\MU x.\; \const{S}( x ) \not\aequiv \MU y.\; \const{S}( \const{S}( y ) )$.
%In the \rn{Acyclic} rule,

@@

However, the presence of quantifiers poses an additional challenge in
the context of (co)datatypes. Quantified formulas can entail an upper bound on
the cardinality of an uninterpreted type $\ty{u}$.
Since it assumes that uninterpreted types have infinite cardinality, the calculus
presented in Section~\ref{sec:the-decision-procedure} is \relax{incomplete}
since it may fail to recognize cases where \rn{Split} and \rn{Single} should be
applied.
This does not %necessarily
impact the correctness of the procedure in this setting,
since %by assumption
the solver is already incomplete for quantified formulas.

Nonetheless, two techniques help increase the precision of the solver.
First, we can apply \rn{Split} to datatype terms whose cardinality depends on the finiteness of
uninterpreted types.
Second, we can conditionally apply \rn{Single} to codatatype terms whose type
potentially has cardinality one.
For example, the codatatype $\ty{stream}_{\vthinspace\ty{u}}$ %, where $\ty{u}$ is uninterpreted.
has cardinality one precisely when $\ty{u}$ has cardinality one.
If there exist two equivalence classes $\ec{s}$ and $\ec{t}$ for this type,
the implementation adds the clause %of the form
$(\exists x\;y\typ{\ty{u}}.\;\, x \not\teq y ) \mathrel\vee s \teq t$ to $F\!$,
which states that either the cardinality of $\ty{u}$ is greater than one or
$s$ must be equal to $t$.

@@

The decision procedure for (co)datatypes is useful both for proving (via
negation, in the refutational style) and for model finding
\cite{reynolds-et-al-2013,ge-de-moura-2009}.
It is in fact vital for finite model finding,
because the acyclicity and uniqueness rules are necessary for solution
soundness, without which the generated models would often be
spurious. For example, given the constraints
\begin{align*}
\const{zeros} & \teq \const{SCons}(0,\:\const{zeros})
& \const{sconst}~n & \teq \const{SCons}(n,\:\const{sconst}~n)
\end{align*}
on streams, the conjecture
$\const{zeros} \teq \const{sconst}~0$
would be ``refuted'' by a spurious countermodel that interprets
\const{zero} and $\const{sconst}~0$ by two distinct
values that both correspond to the term
$\MU s.\; \const{SCons}(0, s)$,
violating uniqueness.
